---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(timeDate)
wd <- 'C:/Users/Micha³.Micha³-PC/Documents/reproducible_research/'
setwd(wd)

### LOADING AND PROCESSING DATA ###
rawData <- read.csv(paste(wd, 'activity.csv', sep=''), header = TRUE, na.strings = "NA")
data <- filter(rawData, steps != 'NA')
data$date <- as.Date(data$date)
head(data, n=10)
```



## What is mean total number of steps taken per day?

```{r message=FALSE, warning=FALSE}
# Constructing data - for median I'm omitting 0 as otherwise most median values for days would be 0
sumByDay <- data %>% group_by(date) %>% summarize(total_steps=sum(steps), mean_steps=mean(steps), median_steps=median(steps[steps>0]))
head(sumByDay)
```

The total steps by day

```{r fig.height=4}
#total
hist(sumByDay$total_steps, 
    main = 'Total Steps Per Day',
    xlab = 'Steps Per Day Totals',
    ylab = 'Frequency',
    col = 'blue',
    las = 1)
```

The mean steps by day

```{r fig.height=4}
hist(sumByDay$mean_steps, 
    main = 'Mean Steps Per Day',
    xlab = 'Mean Steps Per Day',
    ylab = 'Frequency',
    col = 'blue',
    las = 1)
```

The median steps by day

```{r fig.height=4}
hist(sumByDay$median_steps, 
    main = 'Median Steps Per Day',
    xlab = 'Median Steps Per Day',
    ylab = 'Frequency',
    col = 'blue',
    las = 1)
```


## What is the average daily activity pattern?

```{r}
### WHAT IS THE AVERAGE DAILY ACTIVITY PATTERN?

sumByInterval <- data %>% group_by(interval) %>% summarize(total_steps=sum(steps), mean_steps=mean(steps))
head(sumByInterval)
```

Average steps in intervals

```{r}
plot(sumByInterval$interval, sumByInterval$mean_steps, type = "l", ylab='Average steps by Interval', xlab='')
```

Maximum steps in an interval

```{r echo=FALSE}
#max steps in time interval
maxStepsByInterval <- 
  filter(sumByInterval, total_steps == max(sumByInterval$total_steps), max_mean = max(sumByInterval$mean_steps))
head(maxStepsByInterval[, 1:3])
```


## Imputing missing values

```{r}
### IMPUTING MISSING VALUES ###
cleanData <- filter(rawData, complete.cases(rawData))

#missing rows
missingRows <- nrow(rawData) - nrow(cleanData)
```

Rows with NAs:

```{r echo=FALSE}
missingRows
```

Below is the code that imputs missing values to the data set:

1. Filtering the original data set for incomplete rows and assigning to a new data frame
2. Replacing the missing steps with the means for intervals (reusing the sumByInterval data frame which was created for time series plot)
3. Pasting the imputed data with the data frame created after NA values were removed from the original data
4. Ordering the data by date and interval

```{r}
#imputing data
inputData <- filter(rawData, !complete.cases(rawData))
inputData$steps <- replace(inputData$steps, inputData$interval %in% sumByInterval$interval, sumByInterval$mean_steps)

finalData <- rbind(cleanData, inputData) %>% arrange( date, interval)
finalData$date <- as.Date(finalData$date)
head(finalData, n=10)
```

Histogram:

```{r}
#histogram
sumByDay2 <- finalData %>% group_by(date) %>% summarize(total_steps=sum(steps), mean_steps=mean(steps), median_steps=median(steps[steps>0]))

hist(sumByDay2$total_steps, 
    main = 'Total Steps Per Day',
    xlab = 'Steps Per Day Totals',
    ylab = 'Frequency',
    col = 'blue',
    las = 1)
```

Comparison before and after imputing data (all data with imputed values vs raw data without NAs):

```{r}

v1 <- c(mean(finalData$steps), median(finalData$steps), sum(finalData$steps))
v1
```

```{r}

v2 <- c(mean(cleanData$steps), median(cleanData$steps), sum(cleanData$steps))
v2
```

Above, we can see, that median and mean are the same but the total steps count is significantly higher. The median is the same, since there are a lot of 0 in the data. The mean is the same, since I used means for imputing data. 

## Are there differences in activity patterns between weekdays and weekends?

```{r}
### ARE THERE DIFFERENCES IN ACTIVITY PATTERNS BETWEEN WEEKDAYS AND WEEKENDS ###

sumByDay3 <- finalData %>% group_by(date) %>% summarize(total_steps=sum(steps), mean_steps=mean(steps), median_steps=median(steps[steps>0])) %>%
  mutate(day_type = ifelse(isWeekday(date, wday=1:5), 'weekday', 'weekend'))


ggplot(sumByDay3, aes(date, mean_steps)) + geom_line() + facet_grid(day_type ~ .) + xlab('') + ylab('Mean steps')
```

